Complementary roles of Theta and Alpha in Multi-Item Associations

### 2020-01-14 manuscript.tex
added code link
tims feedback on response

\addbibresource{references.bib}
\usepackage{longtable}
\usepackage{csvsimple}				% for reading cvs files into tables

@@

\section*{Acknowledgements}\label{sec:acknowledge}
The authors would like to thank research assistants at Boulder, especially Chris Bird, Krystin Corby, Shaina Martis, as well as other undergraduate volunteers for help in collecting the data used in the EEG experiments.  We would also like to thank faculty at Boulder, specifically McKell Carter, Anu Sharma and Erik Willcutt for helpful feedback during manuscript preparation.  The authors have no known conflicts of interest with this work.  This work was partially funded by NSF Research Fellowship grant \#DGE-0707432.

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), and alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declarative memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically, oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggest the encoding and retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the information that is encoded into the underlying representation.

@@

Building off our previous connectionist model of the hippocampus \parencite{KetzMorkondaOReilly13}, referred to here as the Theta-Phase model, we seek to incorporate the ability to attend to particular dimensions of a given input pattern, while not attending to others.  Practically, this requires that each input be described on all possible stimulus dimensions, otherwise the hippocampal model will attempt to `recall' the missing information for those missing dimensions.  For example, one valid cortical representation experienced will contain information about size and shape but never in color, while another equally valid representation will contain information about color and shape, but never size.  The neural substrate supporting these representations needs to be able to represent all three dimension (and potentially many others) simultaneously and independently, but also be able to understand the inherent bias in which dimensions are likely to covary together.

@@

A simple and ecologically valid way to reconcile this discrepancy is a mechanism that would highlight which of the stimulus dimensions are informative, or relevant given the current input.  This would be very similar to a selective attention mechanism.  Building from the evidence in support of an alpha mediated thalamic gating mechanism, we will explore modulation of cortical activity via thalamo-cortical interactions in the alpha band as a mechanism for the Theta-Phase model to gate or modulate specific stimulus dimensions for targeted encoding and retrieval.  The behavioral and physiological consequences of this mechanism will then be explored in a long-term memory paradigm that utilizes multi-element traces, and a bottom-up attentional signal to focus on specific stimulus dimensions.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figs/depNet_edit2.pdf}
        \caption[Thalamic Gating as Implemented in Theta-Phase Model] {  \emph{(A)} Schematic of Theta-Phase model with thalamic gating.  Posterior cortex connects with both layers 2/3 of Entorhinal cortex (EC) via cortical-cortical projects, but also to thalamic relay cells (TRC) via layer 5 bursting cells.  These TRC units maintain their topography into EC layer 4 where they provide a gating signal for EC to accept or suppress the corresponding cortical-cortical projection. Topographic connectivity shown in dashed arrows, and distributed connectivity shown in solid.  \emph{(B)}  Neural network showing TRC implementation as single units for a given cortical area's input and corresponding unit group in EC.  This topography is kept consistent between TRC and EC unit groups.  \emph{(C)} Illustration of how TRC modulates  network activity during cued-recognition.  Here, a particular person (labeled Per) pattern is being used to cue recognition of an associated animal (labeled Ani) pattern, as indicated by the labels on posterior cortex unit groups.  This input pattern from posterior cortex influences ECin and TRC cells releasing the thalamic suppression in unit groups where there is sufficient activation (i.e. the person and animal unit groups).  The activation from ECin flows up to the hippocampus where missing information can be retrieved, and comes back to ECout through connections from CA1.  ECout's activation influences both ECout and the TRC activations, using the hippocampally retrieved information to fill in the animal unit group as well as show partial recall of a non-target association of location pattern.  The cortical input topography is maintained between EC and TRC where TRC activation is used to calculate the normalized gating signal shown in the corresponding EC units.  Note the gating signal is not representation specific, but rather multiplied across each of the units of its target unit group. 
        }
     \label{fig:depNet}
\end{figure}

@@

As can be seen in Figure \ref{fig:depNet}C, the TRC influences the retrieval of information from the hippocampus by modulating which unit groups in EC are allowed to become active.  If a sufficient input signal from posterior cortex, or a sufficient amount of retrieved information from the hippocampus pushes activity within a EC unit group to provide enough input to the corresponding TRC unit, it starts a positive feedback loop of patterns that are well learned.  During cued recognition, shown in Figure \ref{fig:depNet}C, the target unit group (animal in this case) from posterior cortex is given a uniform, low-level of activity sufficient to start this positive feedback process without actually providing any information about the ultimate pattern of activity.  Retrieved information from the hippocampus can more easily fill in the specifics of this pattern because of the category of the target associate is provided by the recognition candidates (see Figure \ref{fig:deev}b).  This is in contrast to non-target associates, which in this case is in the location unit group.  Retrieved non-target information is being shown influencing ECout, and thereby starting to influence the corresponding TRC unit.  If this retrieved pattern continues to cohere it may provide enough influence to allow ECin to reflect the same pattern and further strength the retrieval process by input that partial pattern back to the hippocampus.

The mechanism used here can be mapped more generally in Figure \ref{fig:attn_compute}

@@

The model's input patterns, described in more detail Section \ref{sec:HippModelValidation}, were created with a base vocabulary of 100 7x7 unit group patterns with 7 units active and a minimum Hamming distance of 10 between each pattern.  The composite input patterns used to train the model were generated from this set of vocabulary patterns, with 8 unit groups in each input pattern corresponding to 8 hypothetical posterior cortical regions.  These input patterns were conceptualized to have 4 unit groups that each corresponded to one of the various element types of the paired associates used in the Dependent Events paradigm.  For each event, one unit group had a unique pattern of activity that corresponded to `location' representations, another for `people', another for `objects', and another for `animals'.  These patterns were consistent across presentations of a given event, i.e. the same unit group patterns were used across the 3 encoding trials, and 6 retrieval trials for each event. In this way selective attention during encoding and retrieval can be  guided to these specific stimulus dimensions of the input pattern, as shown in Figure \ref{fig:depNet}C where two element types are being presented and the corresponding attentional modulation signals are being shown inhibiting the non-attended dimensions.  In addition to these 4 unit groups corresponding to the different event element types, 4 incidental unit groups also maintained the same representation across event presentations.  This was conceptualized as other contextual elements of the event participants imagined, and likely aided in the model's ability to recall the associations (although simulations without these contextual elements were not explored).

@@

To accurately compare condition differences in Dependency between open and closed-loop events its necessary to control for accuracy differences.  For example, if all paired associates are correctly retrieved it's unclear if there is inherent Dependency between event elements or if each association was memorized independently.   To address this, an `Independent' model that reflects the expected Dependency irrespective of the underlying event structure can be calculated to compare with the Dependency inherent in the event structures. For a given model random weight initialization, the mean proportion of correct retrievals for a given element type, say $B$, (over all events) when cued by a given element type, $A$ , can be denoted as $P_{AB}$. This accuracy can be used to create contingency tables for an Independent model where the probability of correctly or incorrectly retrieving two associations is simply the product of the mean probability for each association (i.e., the proportion of events for which both $B$ and $C$ were correctly retrieved when cued by $A$ would be $P_{AB}  P_{AC}$, while the proportion where $B$ was correctly retrieved but $C$ was not would be $P_{AB}(1-P_{AC})$).  A example contingency tables for the Independent model are show in Table \ref{tab:depInd}.

@@

\begin{figure}
	\includegraphics[width=.5\textwidth]{figs/deevTask.pdf}
	\caption[EEG Dependent Events Task]{
		Example stimuli and timing for Dependent Events paradigm adapted for EEG.  All trials had a uniform random inter-trial-interval (ITI) between 0.5s and 1.5s.  Timing was response driving for both cued recognition and confidence judgements.  After 6s a red warning message appeared directing the participant to make their selection.  Feedback was giving by turning grey the choice that was not selected during cued recognition and confidence judgements.  
	}
	\label{fig:deevTask}
\end{figure}

@@

First results are presented for the new proposed model with the augmented thalamic attentional mechanisms, cursorily referred to as the attention augmented model.  Then a comparison of these results with the original Theta-Phase model which has no attentional modulation is made to illustrate the difference in performance with the additional attentional modulation.

Overall performance in the attention augmented model was good for both closed and open-loop events.  Accuracy of cued recall in the network was well above chance (1 of out 100 possible unit group patterns used to determine Name Error, or more conservatively 1 out 36 possible patterns used in the training) in all cue-target types for both conditions, shown in Figure \ref{fig:netDep}A.  This assures that the model was learning the paired associates patterns sufficiently in both conditions, and performing as expected. Closed-loop events did show a higher average accuracy compared to open-loop events across network runs ($\mu=0.14\pm0.06$ 95\% confidence), with the largest difference being when the object or animal elements were the retrieval target, similar to results from \textcite{HornerBisbyBushEtAl15}.

@@

\subsection{Condition Differences in Dependency}
As shown in Figure \ref{fig:netDep}B, Dependency was significantly greater than the Independent model for closed-loop events (paired difference within model runs: $\mu=0.07\pm0.04$ 95\% confidence, $t(7)=3.76,p<0.01$), and not different for open-loop events ($\mu=0.02\pm0.03, t(7)=1.82,p=0.1$).  The interaction between condition differences and Independent model was also significant ($t(7)=2.84,p<0.05$), with the difference between closed-loop Dependency and its Independent model being greater than the difference between open-loop Dependency and its Independent model.

@@

\begin{figure}
    \begin{center}
	\includegraphics[width=\textwidth]{figs/depNonTarg_edit2.pdf}
    \end{center}
      \caption[Non Target Reactivation]{
       \emph{(A)} Name Error accuracy for non-target reactivation in open-loop and closed-loop events.  The horizontal axis labels correspond to the cue and target respectively, and the non-target element is the one not listed in the label.  For example `loc-per' indicates a location cue and person target trial, and therefore the non-target element is either object or animal depending on the encoding pairs for the given event.  The 'oth' label corresponds to either the object or animal element again depending on the corresponding encoding pairs for the given event.  Note, non-target accuracy was not assessed for object or animal elements as cues, as open-loop events have no corresponding non-target element in these cases, and would be a biased comparison with closed-loop events which do. Black dots indicate individual network runs with different training orders (see section \ref{sec:TestProc}), red bars are closed-loop and blue bars are open-loop, dash lines are means across all cue-targ trial types, and error bars are 95\% confidence intervals. \emph{(B)} Example cued retrieval trials from open-loop (top) and closed-loop (bottom) events.  Both show a retrieval trial with a person cue (unit group labeled Per), and a animal target (Ani), and a location non-target (Loc).  TRC shows bottom-up (person and animal unit groups) as well as hippocampal driven (Location unit group) modulation of thalamic attention.  Predicted pattern of oscillatory power for corresponding open and closed loop trials, shown where the higher the TRC activity the lower the expected Alpha power; conversely the lower the higher the hippocampally induced EC activity the higher the Theta power.
	}
     \label{fig:netNonTarg}
\end{figure}

\subsection{Condition Differences in Non-Target Reactivation}
As shown in Figure \ref{fig:netNonTarg}A, successful non-target reactivation, as assessed through Name Error, was significantly greater in the closed-loop condition compared to the open-loop.  Interestingly this difference was most pronounced in the retrieval trials in which the `other' (or object/animal) stimulus dimension was the target of the cued retrieval.  This implies that in open-loop trials, the network was less likely to successfully reactivate  the location or person stimulus elements when attempting to retrieve the animal/object elements.  This can be seen in Figure \ref{fig:netNonTarg}B where cued retrieval is shown for an example open-loop event on top and a closed-loop event on bottom.  Both examples are showing a retrieval trial where the cue element is person (unit group 2, numbering from lower left proceeding up), and the target element is animal (unit group 4), and the non-target element is location (unit group 1).  The relatively lower TRC activation in the non-target element for the open-loop retrieval compared to the closed-loop quantitatively shows the reduced non-target reactivation.  Similarly, the qualitative activity in the closed-loop EC in and out non-target unit group is more complete and closer to the correct pattern as compared to the open-loop activity.

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{figs/deevNetvNull.pdf}
	\end{center}
	\caption[Comparison with original Theta-Phase model]{
	\emph{(A)} Dependency results from original Theta-Phase model.  Here an increase in the difference between data and independent measures of dependency within open-loop trials can be seen compared to augmented model results shown in Figure \ref{fig:netNonTarg}A.
	\emph{(B)} Delta-Dependency: Difference in data and independent measures of dependency when comparing the augmented model with thalamic attention to the original Theta-Phase model.  Here the original model shows a greater difference in dependency in the open-loop stimuli as compared to the augmented model. 
	\emph{(C)} Comparison of Non-target accuracy between the attention augmented model and the original Theta-Phase model.  The attention augmented model shows an increase  Theta-Phase model shows lower accuracy in open-loop animal and object (oth) target stimuli manifest in higher differences in dependency as compared to the thalamic attention model (Figure \ref{fig:netNonTarg}A).  This increase in non-target accuracy illustrates to attention model's ability to represent the association structure of open and closed-loop events in a similar fashion to human subjects.  All error bars are 95\% confidence intervals.
	}
	\label{fig:deevNetvNull}
\end{figure}

\subsection{Comparison with Original Theta-Phase Model}
Results comparing the augmented model with the additional attentional mechanisms to the original Theta Phase model are shown here to illustrate the most direct and meaningful differences in performance.  As shown in Figure \ref{fig:deevNetvNull}, the original Theta-Phase model showed a greater difference between data and independent measures of Dependency within the open-loop stimuli.  This is most directly illustrated in Figure \ref{fig:deevNetvNull}B, where the difference in data vs. independent measures of Dependency are contrasted as the augmented model minus the Theta-Phase.  Here a significant decrease can be seen ($t(14)=3.30$, two sample t-test), which is driven by original Theta-Phase model learning both open and closed-loop stimuli in a similar as illustrated by similar levels of Dependency between the condition types.

Interestingly, both models show no difference in accuracy across cue and target pairings. However, non-target accuracy for animal and object targets (i.e. 'loc-oth' and 'per-oth' in Figure \ref{fig:deevNetvNull}C) showed a significant increase when comparing the augmented model to the Theta-Phase for both open-loop (two sample t-test, location cue: $t(14) = 4.69, p<0.001$, person cue: $t(14)=3.36, p=0.004$) and closed-loop (location cue: $t(14)=4.29, p<0.001$, person cue: $t(14)=3.73, p=0.002$).  Although this non-target increase in accuracy does not directly impact measures of dependency as it can only be inferred through co-activation from particular cue-target trials, it does suggest the mechanism working to manifest the change in dependency measures shown between the original Theta-Phase and augmented models.  Here the Theta-Phase model co-activates unit groups that are not intended as part of the association structure, decreasing non-target accuracy relative to the augmented model.  This over active association in the Theta-Phase model leads to higher differences in data vs. independent measures of Dependency but misrepresents implicit association structure that human subjects are able to capture.  %Similarly, this result provides a novel hypothesis that disruption of the posited thalamic mechanisms in a given biological model would lead to a change in non-target activation.

@@

Critically, there was significant difference between Dependency and the Independent model in the closed-loop condition ($\mu=0.03\pm0.01, t(27)=4.91, p<0.001$), and not in the open-loop condition ($\mu=-0.01\pm0.01, t(27)=-1.78, p=0.09$), and there was also a significant interaction in Dependency by condition, with the difference between closed-loop Dependency and its Independent model being greater then the difference between the open-loop Dependency and its Independent model ($\mu=0.04\pm0.02, t(27)=4.93, p<0.001$).  There were significant differences for accuracy between blocks for closed-loop events (block 1 minus block 2: $\mu=-0.10\pm0.07, t(27)=-2.74, p<0.05$) and Dependency ($\mu=-0.07\pm0.06, t(27)=-2.68, p<0.05$).  Interestingly, these block effects are in the opposite direction as would be predicted from proactive interference, where memory performance progressively gets worse over the blocks.  Similarly, the difference between closed-loop Dependency and its Independent model was significant when considered within each block separately (block 1: $\mu=0.03\pm0.02, t(27)=2.51, p<0.05$; block 2: $\mu=0.04\pm0.02, t(27)=4.49, p<0.001$), however there was no significant interaction in this difference across blocks ($\mu=-0.02\pm0.03, t(27)=-1.32, p=0.2$).

>>>>

\addbibresource{references.bib}
\usepackage{longtable}
\usepackage{csvsimple}				% for reading cvs files into tables
%\usepackage{hyperref}

@@

\section*{Acknowledgements}\label{sec:acknowledge}
The authors would like to thank research assistants at Boulder, especially Chris Bird, Krystin Corby, Shaina Martis, as well as other undergraduate volunteers for help in collecting the data used in the EEG experiments.  We would also like to thank faculty at Boulder, specifically McKell Carter, Anu Sharma and Erik Willcutt for helpful feedback during manuscript preparation.  The authors have no known conflicts of interest with this work.  This work was partially funded by NSF Research Fellowship grant \#DGE-0707432.  Simulation code available at https://github.com/nickketz/deev\_paper/tree/master/Docker
%\href{https://github.com/nickketz/deev_paper/tree/master/Docker}{$https://github.com/nickketz/deev\_paper/tree/master/Docker$}.

\section*{Abstract}\label{sec:abstract}
%The Abstract should be clearly written in 300 words or less and should succinctly state the objectives of the study, experimental design, major observations and conclusions, and their major significance. The abstract should be intelligible to neuroscientists in general and should thus be free of specialized jargon and abbreviations. References should generally not be cited in the abstract, but if they are, the complete citation should be given. Please provide main keywords.
Through the use of neural network models of the hippocampus and surrounding cortex this work proposes a framework for understanding how time frequency signatures measured at the scalp can be used to track declarative memory processes, and make quantitative predictions about how information stored in long term memory is altered by these processes.  The fundamental hypothesis explored here is that neural oscillations in the theta (3-8 Hz), and alpha (8-12 Hz) frequency bands can be tied to specific neural mechanisms supporting declarative memory, and that these oscillatory signatures can be tracked in human scalp EEG recordings to predict behavioral changes in declarative memory processes. Specifically, oscillatory power in the theta band positively correlates with the how much information the hippocampus is reactivating for a given retrieval event, and that power in the alpha band positively correlates with how much information is being inhibited from being retrieved.  We first explore this idea through computational simulations of thalamo-cortical-hippocampal interactions which make a series of predictions tested in a human EEG experiment using the Dependent Events paradigm.  Results largely support the idea that thalamically mediated changes in alpha power during episodic memory encoding and retrieval are related to the flow of information between the neocortex and hippocampus, and that the bandwidth of this information flow tracks with theta power.  Further, this model suggests the encoding and retrieval processes are inter-dependent and that attentional modulation of what gets retrieved from long term memory subsequently impacts the information that is encoded into the underlying representation.

@@

Building off our previous connectionist model of the hippocampus \parencite{KetzMorkondaOReilly13}, referred to through-out the document as the Theta-Phase model, we seek to incorporate the ability to attend to particular dimensions of a given input pattern, while not attending to others.  Practically, this requires that each input be described on all possible stimulus dimensions, otherwise the hippocampal model will attempt to `recall' the missing information for those missing dimensions.  For example, one valid cortical representation experienced will contain information about size and shape but never in color, while another equally valid representation will contain information about color and shape, but never size.  The neural substrate supporting these representations needs to be able to represent all three dimension (and potentially many others) simultaneously and independently, but also be able to understand the inherent bias in which dimensions are likely to covary together.

@@

A simple and ecologically valid way to reconcile this discrepancy is a mechanism that would highlight which of the stimulus dimensions are informative, or relevant given the current input.  This would be very similar to a selective attention mechanism.  Building from the evidence in support of an alpha mediated thalamic gating mechanism, we will develope a new model that explores modulation of cortical activity via thalamo-cortical interactions in the alpha band as a mechanism for the Theta-Phase model to gate or modulate specific stimulus dimensions for targeted encoding and retrieval.  We will refer to this new model as the attentional augmented model through-out the document.  The behavioral and physiological consequences of this mechanism will then be explored in a long-term memory paradigm that utilizes multi-element traces, and a bottom-up attentional signal to focus on specific stimulus dimensions.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figs/depNet_edit2.pdf}
        \caption[Thalamic Gating as Implemented in Theta-Phase Model] {  \emph{(A)} Schematic of Theta-Phase model augmented with attentional modulation.  Posterior cortex connects with both layers 2/3 of Entorhinal cortex (EC) via cortical-cortical projects, but also to thalamic relay cells (TRC) via layer 5 bursting cells.  These TRC units maintain their topography into EC layer 4 where they provide a gating signal for EC to accept or suppress the corresponding cortical-cortical projection. Topographic connectivity shown in dashed arrows, and distributed connectivity shown in solid.  \emph{(B)}  Neural network showing TRC implementation as single units for a given cortical area's input and corresponding unit group in EC.  This topography is kept consistent between TRC and EC unit groups.  \emph{(C)} Illustration of how TRC modulates  network activity during cued-recognition.  Here, a particular person (labeled Per) pattern is being used to cue recognition of an associated animal (labeled Ani) pattern, as indicated by the labels on posterior cortex unit groups.  This input pattern from posterior cortex influences ECin and TRC cells releasing the thalamic suppression in unit groups where there is sufficient activation (i.e. the person and animal unit groups).  The activation from ECin flows up to the hippocampus where missing information can be retrieved, and comes back to ECout through connections from CA1.  ECout's activation influences both ECout and the TRC activations, using the hippocampally retrieved information to fill in the animal unit group as well as show partial recall of a non-target association of location pattern.  The cortical input topography is maintained between EC and TRC where TRC activation is used to calculate the normalized gating signal shown in the corresponding EC units.  Note the gating signal is not representation specific, but rather multiplied across each of the units of its target unit group. 
        }
     \label{fig:depNet}
\end{figure}

@@

As can be seen in Figure \ref{fig:depNet}C, the TRC influences the retrieval of information from the hippocampus by modulating which unit groups in EC are allowed to become active.  If a sufficient input signal from posterior cortex, or a sufficient amount of retrieved information from the hippocampus pushes activity within a EC unit group to provide enough input to the corresponding TRC unit, it starts a positive feedback loop of patterns that are well learned.  During cued recognition, shown in Figure \ref{fig:depNet}C, the target unit group (animal in this case) from posterior cortex is given a uniform, low-level of activity sufficient to start this positive feedback process without actually providing any information about the ultimate pattern of activity.  Retrieved information from the hippocampus can more easily fill in the specifics of this pattern because of the category of the target associate is provided by the recognition candidates (see Figure \ref{fig:deev}b).  This is in contrast to non-target associates, which in this case is in the location unit group.  Retrieved non-target information is being shown influencing ECout, and thereby starting to influence the corresponding TRC unit.  If this retrieved pattern continues to cohere it may provide enough influence to allow ECin to reflect the same pattern and further strength the retrieval process by input that partial pattern back to the hippocampus.  The mechanism used here can be mapped to more general cortical function in Figure \ref{fig:attn_compute}.

@@

The model's input patterns, described in more detail Section \ref{sec:HippModelValidation}, were created with a base vocabulary of 100 7x7 unit group patterns with 7 units active and a minimum Hamming distance of 10 between each pattern.  The composite input patterns used to train the model were generated from this set of vocabulary patterns, with 8 unit groups in each input pattern corresponding to 8 hypothetical posterior cortical regions.  These input patterns were conceptualized to have 4 unit groups that each corresponded to one of the various element types of the paired associates used in the Dependent Events paradigm (i.e. people locations, objects, and animals).  However, the relationship of each unit group to the others is independent, and does not reflect any potential semantic associations that might be present in human subjects.

For each event, one unit group had a unique pattern of activity that corresponded to `location' representations, another for `people', another for `objects', and another for `animals'.  These patterns were consistent across presentations of a given event, i.e. the same unit group patterns were used across the 3 encoding trials, and 6 retrieval trials for each event. In this way selective attention during encoding and retrieval can be  guided to these specific stimulus dimensions of the input pattern, as shown in Figure \ref{fig:depNet}C where two element types are being presented and the corresponding attentional modulation signals are being shown inhibiting the non-attended dimensions.  In addition to these 4 unit groups corresponding to the different event element types, 4 incidental unit groups also maintained the same representation across event presentations.  This was conceptualized as other contextual elements of the event participants imagined, and likely aided in the model's ability to recall the associations (although simulations without these contextual elements were not explored).

@@

To accurately compare condition differences in Dependency between open and closed-loop events its necessary to control for accuracy differences.  For example, if all paired associates are correctly retrieved it's unclear if there is inherent Dependency between event elements or if each association was memorized independently.   To address this, an `Independent' model that reflects the expected Dependency irrespective of the underlying event structure can be calculated to compare with the Dependency inherent in the event structures. For a given model random weight initialization, the mean proportion of correct retrievals for a given element type, say $B$, (over all events) when cued by a given element type, $A$ , can be denoted as $P_{AB}$. This accuracy can be used to create contingency tables for an Independent model where the probability of correctly or incorrectly retrieving two associations is simply the product of the mean probability for each association (i.e., the proportion of events for which both $B$ and $C$ were correctly retrieved when cued by $A$ would be $P_{AB}  P_{AC}$, while the proportion where $B$ was correctly retrieved but $C$ was not would be $P_{AB}(1-P_{AC})$).  Example contingency tables for the Independent model are show in Table \ref{tab:depInd}.

@@

\begin{figure}
	\includegraphics[width=.5\textwidth]{figs/deevTask.pdf}
	\caption[EEG Dependent Events Task]{
		Example stimuli and timing for Dependent Events paradigm adapted for EEG.  All trials had a uniform random inter-trial-interval (ITI) between 0.5s and 1.5s.  Timing was response driven for both cued recognition and confidence judgements.  After 6s a red warning message appeared directing the participant to make their selection.  Feedback was giving by turning grey the choice that was not selected during cued recognition and confidence judgements.  
	}
	\label{fig:deevTask}
\end{figure}

@@

First results are presented for the new proposed model with the augmented thalamic attentional mechanisms, referred to as the attention augmented model.  Then a comparison of these results with the original Theta-Phase model which has no attentional modulation is made to illustrate the difference in performance with the additional attentional modulation.

Overall performance in the attention augmented model was good for both closed and open-loop events.  Accuracy of cued recall in the network was well above chance (1 of out 100 possible unit group patterns used to determine Name Error, or more conservatively 1 out 36 possible patterns used in the training) in all cue-target types for both conditions, shown in Figure \ref{fig:netDep}A.  This assures that the model was learning the paired associates patterns sufficiently in both conditions, and performing as expected. Closed-loop events did show a higher average accuracy compared to open-loop events across network runs ($\mu=0.14\pm0.06$ 95\% confidence), with the largest difference being when the object or animal elements were the retrieval target, similar to results from \textcite{HornerBisbyBushEtAl15}.  This difference is driven by the unique role of animal and object elements in the events' associative structure.  As shown in Figure \ref{fig:deev}, open-loop animal and objects elements have only a single association, compare to other elements which have two.

@@

\subsubsection{Condition Differences in Dependency}
As shown in Figure \ref{fig:netDep}B, Dependency was significantly greater than the Independent model for closed-loop events (paired difference within model runs: $\mu=0.07\pm0.04$ 95\% confidence, $t(7)=3.76,p<0.01$), and not different for open-loop events ($\mu=0.02\pm0.03, t(7)=1.82,p=0.1$).  The interaction between condition differences and Independent model was also significant ($t(7)=2.84,p<0.05$), with the difference between closed-loop Dependency and its Independent model being greater than the difference between open-loop Dependency and its Independent model.

@@

\begin{figure}
    \begin{center}
	\includegraphics[width=\textwidth]{figs/depNonTarg_edit2.pdf}
    \end{center}
      \caption[Non Target Reactivation]{
       \emph{(A)} Name Error accuracy for non-target reactivation in open-loop and closed-loop events.  The horizontal axis labels correspond to the cue and target respectively, and the non-target element is the one not listed in the label.  For example `loc-per' indicates a location cue and person target trial, and therefore the non-target element is either object or animal depending on the encoding pairs for the given event.  The 'oth' label corresponds to either the object or animal element again depending on the corresponding encoding pairs for the given event.  
       %Note, non-target accuracy was not assessed for object or animal elements as cues, as open-loop events have no corresponding non-target element in these cases, and would be a biased comparison with closed-loop events which do. 
       Black dots indicate individual network runs with different training orders (see section \ref{sec:TestProc}), red bars are closed-loop and blue bars are open-loop, dash lines are means across all cue-targ trial types, and error bars are 95\% confidence intervals. \emph{(B)} Example cued retrieval trials from open-loop (top) and closed-loop (bottom) events.  Both show a retrieval trial with a person cue (unit group labeled Per), and an animal target (Ani), and a location non-target (Loc).  TRC shows bottom-up (person and animal unit groups) as well as hippocampal driven (Location unit group) modulation of thalamic attention.  Predicted pattern of oscillatory power for corresponding open and closed loop trials shown, where the higher the TRC activity the lower the expected Alpha power; conversely the higher the hippocampally induced EC activity the higher the Theta power and the lower the TRC activity.
	}
     \label{fig:netNonTarg}
\end{figure}

\subsubsection{Condition Differences in Non-Target Reactivation}
As shown in Figure \ref{fig:netNonTarg}A, successful non-target reactivation, as assessed through Name Error, was significantly greater in the closed-loop condition compared to the open-loop.  Interestingly this difference was most pronounced in the retrieval trials in which the `other' (or object/animal) stimulus dimension was the target of the cued retrieval.  This implies that in open-loop trials, the network was less likely to successfully reactivate  the location or person stimulus elements when attempting to retrieve the animal/object elements.  This can be seen in Figure \ref{fig:netNonTarg}B where cued retrieval is shown for an example open-loop event on top and a closed-loop event on bottom.  Both examples show a retrieval trial where the cue element is person, and the target element is animal, and the non-target element is location.  The relatively lower TRC activation in the non-target element for the open-loop retrieval compared to the closed-loop quantitatively shows the reduced non-target reactivation.  Similarly, the qualitative activity in the closed-loop EC in and out non-target unit group is more complete and closer to the correct pattern as compared to the open-loop activity.

As illustrated in Figure \ref{fig:netNonTarg}B, these difference also illustrate the predicted roles of Theta and Alpha frequency bands.  Here more hippocampal non-target reactivation seen in closed-loop trials would manifest in higher Theta power and lower Alpha, while the comparatively lower reactivation in open-loop trials would manifest in higher Alpha and lower Theta.  Here non-taget TRC activity negatively correlates with Alpha power, and EC non-target activity positively correlates with Theta power.

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{figs/deevNetvNull.pdf}
	\end{center}
	\caption[Comparison with original Theta-Phase model]{
	\emph{(A)} Dependency results from original Theta-Phase model.  Here an increase in the difference between data and independent measures of dependency within open-loop trials can be seen compared to attention augmented model results shown in Figure \ref{fig:netNonTarg}A.
	\emph{(B)} Delta-Dependency: Difference in data and independent measures of dependency when comparing the thalamic attention augmented model to the original Theta-Phase model.  Here the original model shows a greater difference in dependency in the open-loop stimuli as compared to the attention augmented model. 
	\emph{(C)} Comparison of Non-target accuracy between the attention augmented model and the original Theta-Phase model.  The attention augmented model shows an increase in average non-target accuracy compare to the Theta-Phase model.  This difference is driven by the lower non-target accuracy in the Theta-Phase model for open-loop animal and object (oth) target stimuli.  All error bars are 95\% confidence intervals.
	}
	\label{fig:deevNetvNull}
\end{figure}

\subsubsection{Comparison with Original Theta-Phase Model}
Results comparing the attention augmented model with the additional attentional mechanisms to the original Theta Phase model are shown here to illustrate the most direct and meaningful differences in performance.  As shown in Figure \ref{fig:deevNetvNull}, the original Theta-Phase model showed a greater difference between data and independent measures of Dependency within the open-loop stimuli.  This is most directly illustrated in Figure \ref{fig:deevNetvNull}B, where the difference in data vs. independent measures of Dependency are contrasted as the attention augmented model minus the Theta-Phase.  Here a significant decrease can be seen ($t(14)=3.30$, two sample t-test), which is driven by original Theta-Phase model learning both open and closed-loop stimuli in a similar as illustrated by similar levels of Dependency between the condition types.

Interestingly, both models show no difference in accuracy across cue and target pairings. However, non-target accuracy for animal and object targets (i.e. 'loc-oth' and 'per-oth' in Figure \ref{fig:deevNetvNull}C) showed a significant increase when comparing the attention augmented model to the Theta-Phase for both open-loop (two sample t-test, location cue: $t(14) = 4.69, p<0.001$, person cue: $t(14)=3.36, p=0.004$) and closed-loop (location cue: $t(14)=4.29, p<0.001$, person cue: $t(14)=3.73, p=0.002$).  Although this non-target increase in accuracy does not directly impact measures of dependency as it can only be inferred through co-activation from particular cue-target trials, it does suggest the mechanism working to manifest the change in dependency measures shown between the original Theta-Phase and attention augmented models.  Here the Theta-Phase model co-activates unit groups that are not intended as part of the association structure, decreasing non-target accuracy relative to the attention augmented model.  This over active association in the Theta-Phase model leads to higher differences in data vs. independent measures of Dependency but misrepresents implicit association structure that human subjects are able to capture.  %Similarly, this result provides a novel hypothesis that disruption of the posited thalamic mechanisms in a given biological model would lead to a change in non-target activation.

@@

Critically, there was significant difference between Dependency and the Independent model in the closed-loop condition ($\mu=0.03\pm0.01, t(27)=4.91, p<0.001$), and not in the open-loop condition ($\mu=-0.01\pm0.01, t(27)=-1.78, p=0.09$), and there was also a significant interaction in Dependency by condition, with the difference between closed-loop Dependency and its Independent model being greater then the difference between the open-loop Dependency and its Independent model ($\mu=0.04\pm0.02, t(27)=4.93, p<0.001$), see Figure \ref{fig:deevEEGBeh}B.  There were significant differences for accuracy between blocks for closed-loop events (block 1 minus block 2: $\mu=-0.10\pm0.07, t(27)=-2.74, p<0.05$) and Dependency ($\mu=-0.07\pm0.06, t(27)=-2.68, p<0.05$).  Interestingly, these block effects are in the opposite direction as would be predicted from proactive interference, where memory performance progressively gets worse over the blocks.  Similarly, the difference between closed-loop Dependency and its Independent model was significant when considered within each block separately (block 1: $\mu=0.03\pm0.02, t(27)=2.51, p<0.05$; block 2: $\mu=0.04\pm0.02, t(27)=4.49, p<0.001$), however there was no significant interaction in this difference across blocks ($\mu=-0.02\pm0.03, t(27)=-1.32, p=0.2$), see Figure \ref{fig:deevEEGBeh}C.
